---
title: "Predictive Biomarker Analysis: FinHER Trial"
author: "Philippe Boileau"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(here)
library(GEOquery)
library(hgu219.db)
library(genefilter)
library(matrixStats)
library(data.table)
library(survival)
library(ggplot2)
library(ggfortify)
library(sl3)
library(unihtee)
library(knitr)
library(kableExtra)
library(dplyr)
library(pheatmap)
```

# Prepare the Data

Where does the data come from? What kind of cancer? Treatment?

## Load the Data

An `ExpressionSet` object is created from Series GSE47994 data. It contains the
expression and clinical data of 335 early-stage, triple negative breast cancer
patients. Two hundred and two of these patients are classified as HER2+.

```{r load-data, cache=TRUE}
eset <- getGEO(
  GEO = "GSE47994",
  dest = here("analysis/data"),
  GSEMatrix = TRUE,
  getGPL = FALSE
)[[1]]
```

These patients' expression data has already been processed using the Robust
multi-array analysis method, followed by quantile normalization. Probe-level
annotations were obtained from the assay manufacturer's web site.

## Preprocessing 

Only HER2+ patients are retained for this analyis. We also limit our analysis to
the 500 most variable genes that are well-annotated.

```{r filter-data}
## retain only HER2+ positive patients
eset <- eset[, pData(eset)$`her2 status:ch1` == "1"]

## retain only most variable probes
## partial probe removal (uses annotation info)
annotation(eset) <- "hgu219" 
eset <- nsFilter(eset)$eset
probe_variances <- rowVars(exprs(eset))
keep_probes <- rank(-probe_variances) <= 500
eset <- eset[keep_probes]
```

We then replace the current feature names by their human-friendly gene symbol.

```{r get-symbols}
rownames(eset) <- as.vector(unlist(mget(rownames(eset), hgu219SYMBOL)))
```

Finally, we prepate the `data.table` object for eventual use with `unihtee()`.

```{r make-data.table}
## turn expression data into a data.table
exprs_dt <- as.data.table(t(exprs(eset)))

## extract relevant observation-level data
obs_dt <- as.data.table(pData(eset))[,
  `:=`(
    age = as.numeric(`patient age (years):ch1`),
    recurrence = as.numeric(`distant recurrence:ch1`),
    time = as.numeric(`time to distant recurrence (years):ch1`),
    docetaxel = as.numeric(`chemotherapy regimen:ch1` == "Docetaxel"),
    herceptin = as.numeric(`herceptin treatment:ch1`)
  )]
obs_dt <- obs_dt[, c("recurrence", "time", "herceptin", "docetaxel", "age")]
obs_dt$prop_score <- 0.5

## combine the data, and retain only complete cases
dt <- cbind(obs_dt, exprs_dt)
dt <- dt[complete.cases(dt), ]
```


## Exploratory Data Analysis

Plot the outcome of interest. Is the proportional hazards assumption violated?
If not, then we can't consider the penalized cox models here. Discuss
assumptions for RMST-based TEM-VIP assumptions.

We next perform a brief exploratory data analysis. First, we ensure that the
number of patients receiving Herceptin is approximately equal to the number of
patients in the control arm. We further stratify by chemo regimen to assess
balance. The treatment is approximately perfectly balanced across patients, as
are docetaxel versus vinorelbine.

```{r therapy-table}
table("docetaxel" = dt$docetaxel, "herceptin" = dt$herceptin)
```

We now plot the Kaplan Meyer curve for DDFS.

```{r KM}
km_fit <- survfit(Surv(time, recurrence) ~ herceptin, data = dt)
autoplot(km_fit) + 
 labs(x = "\n Survival Time (Years) ", y = "Survival Probabilities \n")
```

It's clear from this plot that there herceptin does not provide a strong
protective effect.

# Predictive Biomarker Discovery

## Apply `unihtee`

Apply `unihtee` to estimate the RMST-based treatment effect modifier variable
importance parameter. Let's use the same library of learners as in the
simulation. Sensitivity analysis with more complicated SL library. Use the TMLE
estimator since it was better in this setting during simulations.

```{r unihtee}
## discretize the data to be in half years, and create a censoring indicator
dt[, `:=`(
  discrete_time = ceiling(time * 2) / 2,
  censoring = 1 - recurrence
)]

## define the confounders and potential treatment effect modifiers
confounders <- c(c("age", "docetaxel"), colnames(exprs_dt))
modifiers <- colnames(exprs_dt)

## build a super learner
lrnr_interactions <- Lrnr_define_interactions$new(
  lapply(modifiers, function(mod) c("herceptin", mod))
)
lrnr_lasso_haz <- make_learner(
  Pipeline, lrnr_interactions, Lrnr_glmnet$new()
)
lrnr_lasso_cens <- Lrnr_glmnet$new()
lrnr_ranger <- Lrnr_ranger$new()
lrnr_xgboost <- Lrnr_xgboost$new()

## define the discrete super learners
bin_metalearner <- make_learner(
  Lrnr_solnp, metalearner_logistic_binomial, loss_squared_error
)
lrnr_sl_haz <- Lrnr_sl$new(
  learners = list(
    lrnr_lasso_haz, lrnr_ranger, lrnr_xgboost
  ),
  metalearner = bin_metalearner
)
lrnr_sl_cens <- Lrnr_sl$new(
  learners = list(
    lrnr_lasso_cens, lrnr_ranger, lrnr_xgboost
  ),
  metalearner = bin_metalearner
)

set.seed(514)
results_tmle <- unihtee(
  data = dt,
  confounders = confounders,
  modifiers = modifiers,
  exposure = "herceptin",
  outcome = "discrete_time",
  censoring = "censoring",
  time_cutoff = 6,
  outcome_type = "time-to-event",
  effect = "absolute",
  estimator = "tmle",
  prop_score_values = "prop_score",
  failure_hazard_estimator = lrnr_sl_haz,
  censoring_hazard_estimator = lrnr_sl_cens
)
```

## Empirical Cumulative Distribution Function of P-Values

```{r ecdf-p-values}
results_tmle %>%
  ggplot(aes(p_value)) +
  stat_ecdf(pad = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 1), linetype = 2) +
  ylim(0, 1) +
  xlim(0, 1) +
  xlab("Nominal P-Values") +
  ylab("Cumulative Probability") +
  theme_bw()

ggsave(
  "tmle-ecdf.jpeg",
  path = here("analysis/results"),
  scale = 1.2,
  height = 3,
  width = 5,
  dpi = "retina"
)
```

## Table of Statistically Significant Biomarkers

```{r significant-genes}
results_tmle[p_value_fdr <= 0.05] %>%
  kbl(digits = 3) %>%
  kable_styling
```

## Volcano Plot

```{r volcano}
results_tmle %>%
  mutate(
    neg_log10_p_value = -log10(p_value),
    biomarker_cat = case_when(
      p_value_fdr <= 0.05 & abs(estimate) > 0.05 ~ "hit",
      p_value_fdr > 0.05 & abs(estimate) > 0.05 ~ "interesting",
      abs(estimate) <= 0.1 ~ "uninteresting"
    ),
    biomarker_cat = factor(biomarker_cat,
      levels = c("hit", "interesting", "uninteresting")
    )
  ) %>%
  ggplot(aes(
    x = estimate, y = neg_log10_p_value, colour = biomarker_cat
  )) +
    geom_point() +
    geom_hline(yintercept = -log10(0.024), linetype = 2, alpha = 0.3) +
    geom_vline(xintercept = 0.05, linetype = 2, alpha = 0.3) +
    geom_vline(xintercept = -0.05, linetype = 2, alpha = 0.3) +
    scale_color_manual(
      values = c("hit" = "red", "interesting" = "orange", "uninteresting" = "yellow")
    ) +
    xlab("TEM-VIP Estimates") +
    ylab("-log10(P-Value)") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
  "tmle-volcano.jpeg",
  path = here("analysis/results"),
  scale = 1.2,
  height = 3,
  width = 5,
  dpi = "retina"
)
```

## Predicted Predictive Biomarkers' Expression Heatmap

A total of `r nrow(results_tmle[(abs(estimate) >= 0.05 & p_value_fdr <= 0.05)])`
biomarkers are deemed to modify the effect of treatment in a clinically
meaningful way.

```{r heatmap, fig.width=3}
predictive_biomarkers <- results_tmle[
  (abs(estimate) >= 0.05 & p_value_fdr <= 0.05), modifier]

## save the heatmap
pheatmap(
  mat = as.matrix(exprs_dt[, ..predictive_biomarkers]),
  width = 8,
  height = 12,
  filename = here("analysis/results/tmle-heatmap.jpeg"),
  border_color = NA,
  show_rownames = FALSE,
  show_colnames = FALSE
)
knitr::include_graphics(
  here("analysis/results/tmle-heatmap.jpeg"),
  error = FALSE
)
```
